---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mapscvi

<!-- badges: start -->
<!-- badges: end -->

Map single cell expression data in a seurat object into reference scvi latent space and reference umap based on seurat.

## Installation

Install mapscvi using:

``` r
devtools::install_github("lsteuernagel/mapscvi")
```

## Important links:

In order to use the package python and scvi >= 0.15, as well as R and Seurat > 4.0.0 are required.

A docker image which comes with a compatible R, Seurat v4, pytorch and scvi installation can be found [here](https://hub.docker.com/r/lsteuernagel/r_scvi_docker_v2)

The hypoMap model is included in this paper and can be found [here](inst/extdata/models/hypoMap_harmonized_scVI_model/)

The full hypoMap can be downloaded from this [repository](https://www.repository.cam.ac.uk/handle/1810/340518)
The wrapper function defaults to a smaller object (that can be shipped with the package) for reference_seurat (mapscvi::reference_hypoMap_downsample).

Please see [this folder](inst/python_notebook/) for a simple python example notebook if you don't want to work through R.

## Example

This package allows to embed new single-cell data stored in Seurat objects into HypoMap.

The functions used to do this can also be used to embed data into other models.

An example workflow that embeds the Romanov et al. smart-seq dataset into the HypoMap:

```{r, message=FALSE, warning=FALSE}
library(mapscvi)
library(Seurat)
```

We can use the example data which is a Seurat object.

```{r}
query_romanov
```

The test data does not contain any dimensional reductions for visualization or annotation.

```{r}
names(query_romanov@reductions)
```


#### Map the data onto the HypoMap

To map data to the reference HypoMap, we use the HypoMap model and a reduced reference seurat object with only relevant information and fewer cell.
The model and reference seurat for neighbor search and umap construction will automatically be used when setting the reference_mode to "hypoMap_neurons" or  "hypoMap_full" in the wrapper function below.

```{r echo=TRUE, message=FALSE, warning=FALSE, results = 'hide'}
query_romanov = mapscvi::map_new_seurat_hypoMap(mapscvi::query_romanov,suffix="query_romanov",label_col = "C66_named",max_epochs=20)
```

After successful mapping the objects contains the scvi and umap reductions:

```{r}
names(query_romanov@reductions)
```


#### Plotting results:

We can take a look at the top clusters that were found in the query:

```{r}
head(sort(table(query_romanov@meta.data$predicted),decreasing = TRUE),n = 10)
```

The package provides plotting functions to visualize the query cells on the reference:

We can plot query and reference side by side. Here we set labelonplot to False to prevent cluster labels from being plotted and we use the object 'reference_hypoMap_full' that is automatically loaded when using the wrapper above.

```{r}
plot_query_labels(query_seura_object=query_romanov,reference_seurat=mapscvi::reference_hypoMap_downsample,label_col="C66_named",overlay = FALSE,labelonplot = FALSE)
```

Overlay them query over the reference. The overlay parameters allow to change the behavior of query points. We can use the Seurat::DimPlot parameters to further adjust the plots. E.g. by decreasing the size of the labels.

```{r}
plot_query_labels(query_seura_object=query_romanov,reference_seurat=mapscvi::reference_hypoMap_downsample,label_col="C66_named",overlay = TRUE,query_pt_size = 0.4,labelonplot = FALSE,label.size=1)
```

The mapping also returns a prediction probability based on the similarity to the the neighbors in there reference which an indicate how well different cells mapped to their assigned celltypes.

```{r}
FeaturePlot(query_romanov,features = "prediction_probability")+NoAxes()
```


## Detailed walkthrough

This section contains a more detailed walkthrough of the functions that are executed when calling 'map_new_seurat_hypoMap'.

For this we will embed the mouse LaMannoBrainData (available via the scRNAseq package) in the HypoMap.

```{r, echo=TRUE, message=FALSE, warning=FALSE, results = 'hide'}
lamanno_seurat
```

The prepare_query function is able to load Seurat, SingleCellExperiment or matrix objects for mapping.

```{r message=FALSE, warning=FALSE}
lamanno_seurat_object = prepare_query(lamanno_seurat,suffix="lamanno_da",normalize=TRUE)
```

This new seurat object is compatible with the downstream functions for mapping the data.

Next, predict_query can be used to embed the query data into the latent space of scvi. We have to specify a model path and the number of epochs for training during mapping (10-20 should be sufficient.).

The mapscvi package comes with two models for the neuron and full hypothalamus map in the extdata which can be found using system.file as below.

```{r message=FALSE, warning=FALSE, results = 'hide'}
model_path = paste0(system.file('extdata/models/hypoMap_harmonized_scVI_model/"', package = 'mapscvi'),"/")
max_epochs = 20
lamanno_seurat_object = predict_query(lamanno_seurat_object,model_path = system.file("extdata/models/hypoMap_harmonized_scVI_model/", package = 'mapscvi'),max_epochs = max_epochs)
```

```{r}
names(lamanno_seurat_object@reductions)
```

The scvi reduction is a pca-like low dimensional space that can be used to embed the data into the same UMAP as the reference object. This requires an existing UMAP model in the reference Seurat object that was calculated based on the same scvi latent space.

We use the mapscvi::reference_hypoMap_downsample object that comes with the package and contains for each cell the scvi and umap reduction values as well as metadata.

```{r}
mapscvi::reference_hypoMap_downsample
```
Then we can calculate nearest neighbors and UMAP based on the reference. Additionally we can project labels (any categorical metadata column from the reference) using the same nearest neighbor graph. This helps with consistency between label propagation and UMAP. However there might be more accurate ways to propagate labels using other classifiers such as a random forest or scANVI.

To propagate labels with the project_query function we can provide a vector of the same length as the reference cells (and same order!). Preferably this is a column from the metadata of the reference seurat object.

```{r, message=FALSE, warning=FALSE, results = 'hide'}
cluster_labels = mapscvi::reference_hypoMap_downsample@meta.data$C66_named
reference_reduction = "scvi"
lamanno_seurat_object = project_query(query_seurat_object = lamanno_seurat_object,
                                      reference_map_reduc = mapscvi::reference_hypoMap_downsample@reductions[[reference_reduction]],
                                      reference_map_umap = mapscvi::reference_hypoMap_downsample@reductions[[paste0("umap_",reference_reduction)]],
                                      query_reduction = "scvi",
                                      label_vec =cluster_labels)
```

This can then be used to plot the results side-by side:

```{r}
plot_query_labels(query_seura_object=lamanno_seurat_object,reference_seurat=mapscvi::reference_hypoMap_downsample,label_col="C66_named",overlay = FALSE,labelonplot = FALSE)
```

Or with overlay:

```{r}
plot_query_labels(query_seura_object=lamanno_seurat_object,reference_seurat=mapscvi::reference_hypoMap_downsample,label_col="C66_named",overlay = TRUE,query_pt_size = 0.4,labelonplot = FALSE,label.size=1)
```
This obviously didn't work out very well.

Most cells were mapped to the same celltype in the middle of the UMAP, which is likely a result of missing 'true' neighbors:

```{r}
head(sort(table(lamanno_seurat_object@meta.data$predicted),decreasing = TRUE),n = 10)
```
This is not very suprising because the datasets does not contain hypothalamic neurons, while the nearest neighbor algorithm will still try to find neighbors in the reference.
Cholinergic neurons are moved to the hypothalamic cholinergic neurons.

#### Quality of mapping results:

We can use various quality metrics to see how the projection of the new data worked.

The prediction_probability is mostly helpful in determining whether a celltype was unambiguously mapped. 

```{r}
FeaturePlot(lamanno_seurat_object,features = "prediction_probability")+NoAxes()
```

## Conclusion

This vignette demonstrated the use of the mapscvi package on two example datasets to embed unannotated single cell data in the hypothalamus reference map.
Based on the detailed walkthrough section it is also possible to adapt the package for custom models and objects. 




