---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mapscvi

<!-- badges: start -->
<!-- badges: end -->

Map single cell expression data in a seurat object into reference scvi latent space and reference umap based on seurat.

## Installation

Install mapscvi using:

``` r
devtools::install_github("lsteuernagel/mapscvi")
```

In order to use the package python and scvi >= 0.8, as well as R and Seurat > 4.0.0 are required.

A docker image which comes with a compatible R, Seurat v4, pytorch and scvi installation can be found here: https://hub.docker.com/r/lsteuernagel/r_scvi 

## Example

This package allows to embed new single-cell data stored in Seurat objects into HypoMap.

The functions used to do this can also be used to embed data into other models.

An example workflow that embeds the Romanov et al. smart-seq dataset into the HypoMap:

```{r, message=FALSE, warning=FALSE}
library(mapscvi)
```

We can use the example data which is a Seurat object.

```{r}
query_romanov
```

The test data does not contain any dimensional reductions for visualization or annotation.

```{r}
names(query_romanov@reductions)
```

#### Obtain the reference HypoMap

At the momment the reference seurat object cannot be distributed directly as part of the package.

Load it directly (e.g. after download):

```{r echo=TRUE, message=FALSE, warning=FALSE, results = 'hide'}
neuron_map_seurat = readRDS("/beegfs/scratch/bruening_scratch/lsteuernagel/data/hypoMap/hypoMap_objects/hypothalamus_neurons_reference.rds")
```


#### Map the data onto the HypoMap

To map data to the reference HypoMap, we use the HypoMap model and a reduced reference seurat object with only relevant information and fewer cell.
The model will automatically be used when setting the reference_mode to "hypoMap_neurons" or  "hypoMap_full" in the wrapper function below.
However we still have to provide the reference object loaded above!

```{r echo=TRUE, message=FALSE, warning=FALSE, results = 'hide'}
query_romanov_neurons = map_new_seurat_hypoMap(mapscvi::query_romanov,reference_mode = "hypoMap_neurons",reference_seurat = neuron_map_seurat, suffix="query_romanov_neurons",max_epochs=20)
names(query_romanov@reductions)
```

We can take a look at the top clusters that were found in the query:

```{r}
head(sort(table(query_romanov_neurons@meta.data$predicted),decreasing = TRUE),n = 10)
```

The package provides plotting functions to visualize the query cells on the reference:

We can plot query and reference side by side. Here we set labelonplot to False to prevent cluster labels from being plotted and we use the object 'reference_hypoMap_full' that is automatically loaded when using the wrapper above.

```{r}
plot_query_labels(query_seura_object=query_romanov_neurons,reference_seurat=neuron_map_seurat,label_col="K169_named",overlay = FALSE,labelonplot = FALSE)
```

Overlay them query over the reference. The overlay parameters allow to change the behavior of query points. We can use the Seurat::DimPlot parameters to further adjust the plots. E.g. by decreasing the size of the labels.

```{r}
plot_query_labels(query_seura_object=query_romanov_neurons,reference_seurat=neuron_map_seurat,label_col="K169_named",overlay = TRUE,query_pt_size = 0.4,labelonplot = FALSE,label.size=1)
```

## Detailed walkthrough

This section contains a more detailed walkthrough of the functions that are executed when calling 'map_new_seurat_hypoMap'.

For this we will embed the mouse LaMannoBrainData stored in a SingleCellExperiment (available via the scRNAseq package) in the HypoMap.

```{r, echo=TRUE, message=FALSE, warning=FALSE, results = 'hide'}
sce_lamanno_da
```
The prepare_query function is able to load Seurat, SingleCellExperiment or matrix objects for mapping.

Here we are using the prepare_query_hypoMap that extends the more generic prepare_query function with code to to automatically add metadata variables that are required for mapping to the HypoMap model. The function can also normalize the data, however for the scvi mapping raw counts are required!

```{r message=FALSE, warning=FALSE}
lamanno_seurat_object = prepare_query_hypoMap(sce_lamanno_da,suffix="lamanno_da",normalize=TRUE)
lamanno_seurat_object
```

This new seurat object is compatible with the downstream functions for mapping the data.

Next predict_query can be used to embed the query data into the latent space of scvi. We have to specify a model path and the number of epochs for training during mapping (10-20 should be sufficient. TODO: test this!).

```{r message=FALSE, warning=FALSE, results = 'hide'}
model_path = paste0(system.file('extdata/models/hypothalamus_neurons_reference_model', package = 'mapscvi'),"/")
max_epochs = 20
lamanno_seurat_object = predict_query(lamanno_seurat_object,model_path,max_epochs = max_epochs)
```

```{r}
names(lamanno_seurat_object@reductions)
```

The scvi reduction is a pca-like low dimensional space that can be used to embed the data into the same UMAP as the reference object. This requires an existing UMAP model in the reference Seurat object that was calculated based on the same scvi latent space.

We use the neuron_map_seurat object loaded in the first section!

Then we can calculate nearest neighbors and UMAP based on the reference. Additionally we can project labels (any categorical metadata column from the reference) using the same nearest neighbor graph. This helps with consistency between label propagation and UMAP. However there might be more accurate ways to propagate labels using other classifiers such as a random forest or scANVI.

To propagate labels with the project_query function we can provide a vector of the same length as the reference cells (and same order!). Preferably this is a column from the metadata of the reference seurat object.

```{r message=TRUE, warning=TRUE, results = 'hide'}
cluster_labels = neuron_map_seurat@meta.data$K169_named
reference_reduction = "scvi"
lamanno_seurat_object = project_query(query_seurat_object = lamanno_seurat_object,
                                      reference_map_reduc = neuron_map_seurat@reductions[[reference_reduction]],
                                      reference_map_umap = neuron_map_seurat@reductions[[paste0("umap_",reference_reduction)]],
                                      query_reduction = "scvi",
                                      label_vec =cluster_labels)
```

This can then be used to plot the results side-by side:

```{r}
plot_query_labels(query_seura_object=lamanno_seurat_object,reference_seurat=neuron_map_seurat,label_col="K169_named",overlay = FALSE,labelonplot = FALSE)
```

Or with overlay:

```{r}
plot_query_labels(query_seura_object=lamanno_seurat_object,reference_seurat=neuron_map_seurat,label_col="K169_named",overlay = TRUE,query_pt_size = 0.4,labelonplot = FALSE,label.size=1)
```
This obviously didn't work out very well.

Most cells were mapped to the same celltype in the middle of the UMAP, which is likely a result of missing 'true' neighbors:

```{r}
head(sort(table(lamanno_seurat_object@meta.data$predicted),decreasing = TRUE),n = 10)
```

TODO: Add some indicators for quality of mapping.




